\documentclass[aspectratio=169,10pt]{beamer}
\usetheme{metropolis}
\usepackage{booktabs}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{xcolor}

\definecolor{accent}{HTML}{e74c3c}
\definecolor{good}{HTML}{27ae60}
\definecolor{bad}{HTML}{e74c3c}
\setbeamercolor{alerted text}{fg=accent}

\title{Entropy-Guided Token Importance for PPO}
\subtitle{Adaptive Intensity Weighting for Math Reasoning}
\author{Gurusha Juneja}
\date{March 2026}

\begin{document}

\maketitle

% ============================================================
% SLIDE 1: The Problem
% ============================================================
\begin{frame}{The Problem: Not All Tokens Are Equal in RLHF}

Standard PPO treats every token equally in the policy gradient:
\[
\mathcal{L}_{\text{PPO}} = \frac{1}{T}\sum_{t=1}^{T} \min\!\bigl(r_t \hat{A}_t,\; \text{clip}(r_t)\,\hat{A}_t\bigr)
\]

\textbf{But tokens contribute unequally to learning:}
\begin{itemize}
    \item \textit{``Let $x$ = \alert{3 + 4 = 7}. Therefore the answer is \alert{\textbackslash boxed\{7\}}''} --- the reasoning step matters
    \item \textit{``The answer is''} --- boilerplate, no learning signal
    \item Function words (``the'', ``is'', ``let'') are noise in the gradient
\end{itemize}

\vspace{0.3cm}
\textbf{Question:} Can we weight tokens by importance to get \textit{faster learning} and \textit{lower KL divergence}?

\end{frame}

% ============================================================
% SLIDE 2: TI-DPO Background
% ============================================================
\begin{frame}{Prior Work: TI-DPO (Yang et al., 2025)}

Token-Importance guided DPO weights the preference loss per-token:
\[
\mathcal{L}_{\text{TI-DPO}} = -\log\sigma\!\Bigl(\beta\sum_t \alert{w(t)}\bigl[\log\pi(a_t^w|s_t) - \log\pi(a_t^\ell|s_t)\bigr]\Bigr)
\]

\vspace{0.3cm}

\textbf{Their importance scoring methods:}

\vspace{0.2cm}
\begin{table}
\small
\begin{tabular}{lll}
\toprule
Method & Signal & Cost \\
\midrule
Gradient & $\|\nabla_{\text{emb}}\|_1$ per token & 1 backward \\
Attention & Mean attention received & 1 forward \\
Hybrid & Gradient + Gaussian prior & 1 backward + 1 forward \\
TD Error & $|V(s_t) - V(s_{t+1}) - r_t|$ & Free \\
\bottomrule
\end{tabular}
\end{table}

\vspace{0.2cm}
Works well for DPO: +2--5\% on preference benchmarks.

\end{frame}

% ============================================================
% SLIDE 3: Our first step --- port to PPO
% ============================================================
\begin{frame}{Our First Step: Port TI-DPO Methods to PPO}

We use the \textbf{same importance scores} from TI-DPO as per-token weights in the PPO loss:
\[
\mathcal{L} = \frac{1}{T}\sum_t \alert{w(t)} \cdot \min\!\bigl(r_t \hat{A}_t,\;\text{clip}(r_t)\hat{A}_t\bigr)
\]

where $w(t)$ is one of the paper's scores: \textbf{gradient norm} $\|\nabla_{\text{emb}_t}\|_1$, \textbf{attention} received, or the \textbf{hybrid} (gradient + Gaussian prior).

\vspace{0.3cm}
\textbf{But there is a key difference between DPO and PPO:}

\begin{columns}
\begin{column}{0.48\textwidth}
\textbf{DPO (offline)}
\begin{itemize}
    \item Paired (chosen, rejected) sequences
    \item Single pass over data
    \item These scores measure ``which tokens differ between chosen/rejected?''
\end{itemize}
\end{column}
\begin{column}{0.48\textwidth}
\textbf{PPO (online)}
\begin{itemize}
    \item Single sequence + scalar reward
    \item Hundreds of gradient steps
    \item No contrastive pair $\Rightarrow$ gradient/attention scores have no clear meaning
\end{itemize}
\end{column}
\end{columns}

\end{frame}

% ============================================================
% SLIDE 4: Results after Phase 1
% ============================================================
\begin{frame}{Phase 1 Results: Direct Port Fails}

\textit{\scriptsize Synthetic benchmark: GPT-2 124M + LoRA, sentiment reward, 150 episodes}
\begin{table}
\begin{tabular}{lccc}
\toprule
Method & Reward (Q4) & KL (Q4) & KL-Efficiency \\
\midrule
PPO baseline (uniform) & \textbf{0.850} & 0.043 & 19.9 \\
Paper: Hybrid+Triplet & 0.779 & 0.024 & 32.7 \\
\bottomrule
\end{tabular}
\end{table}

\vspace{0.3cm}
The hybrid method is KL-efficient (low divergence) but achieves \textbf{8.4\% less reward} than vanilla PPO.

\vspace{0.3cm}
\textbf{Takeaway:} We need importance signals that are native to the PPO setting, not borrowed from DPO.

\end{frame}

% ============================================================
% SLIDE 5: Entropy as importance
% ============================================================
\begin{frame}{Our Idea: Use Entropy as a Token Importance Signal}

\textbf{Per-token policy entropy} measures the model's uncertainty at each position:
\[
H\bigl(\pi(\cdot | s_t)\bigr) = -\sum_v \pi(v|s_t) \log \pi(v|s_t)
\]

We normalize and use it as an importance weight in the PPO loss:
\[
w(t) = \frac{H(t)}{\frac{1}{T}\sum_{t'} H(t')} \qquad \Longrightarrow \qquad \mathcal{L} = \frac{1}{T}\sum_t \alert{w(t)} \cdot \min(r_t \hat{A}_t,\;\text{clip}(r_t)\hat{A}_t)
\]

\textbf{Why entropy?}
\begin{itemize}
    \item \textbf{High entropy} = model sees multiple plausible next tokens = \textcolor{good}{decision point, room to learn}
    \item \textbf{Low entropy} = model is confident, one dominant choice = less to gain from updates
    \item \textbf{Free:} computed from existing logits, zero extra cost
    \item \textbf{Content-aware:} depends on \textit{what} the token means, not \textit{where} it is
\end{itemize}

\end{frame}

% ============================================================
% SLIDE 6: Entropy intuition
% ============================================================
\begin{frame}{Entropy Intuition: Where Should the Model Focus?}

\textbf{Example:} ``Solve: What is $15 \times 7$?''

\vspace{0.2cm}
\begin{center}
\begin{tabular}{lcl}
\toprule
Token & $H(\pi)$ & Interpretation \\
\midrule
``Let'' & 0.2 & Boilerplate --- model is sure \\
``me'' & 0.1 & Continuation is obvious \\
``calculate'' & 0.8 & Multiple valid reasoning strategies \\
``15'' & 0.3 & Copying from the problem \\
``$\times$'' & 0.1 & Determined by context \\
``='' & \alert{1.4} & \textbf{Critical: model must compute the answer!} \\
``105'' & \alert{1.2} & \textbf{The actual answer --- high uncertainty} \\
``.'' & 0.1 & Punctuation \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.2cm}
Entropy naturally identifies \textbf{decision points} --- tokens where the model's choice determines correctness. PPO's gradient is most useful precisely at these positions.
\end{frame}

% ============================================================
% SLIDE 6b: Why not just rely on grad log pi?
% ============================================================
\begin{frame}{Doesn't $\nabla\log\pi$ Already Know Which Tokens Matter?}

\begin{center}
\begin{tabular}{lccc}
\toprule
Token type & $\|\nabla\log\pi\|$ & $H(\pi)$ & Useful to update? \\
\midrule
Confident \& correct (``the'') & small & low & No \\
Confident \& \textbf{wrong} (``52'' not ``105'') & small & low & Yes, but gradient is small! \\
\alert{Uncertain (``='' before answer)} & medium & \alert{high} & \alert{Yes --- most leverage} \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.3cm}
$\nabla\log\pi$ reflects \textit{how much the output would change} --- not \textit{where a change would help}.

\vspace{0.3cm}
Entropy adds what the gradient lacks: \textbf{which tokens are at a decision boundary} where a small update can tip the model toward the right answer.

\end{frame}

% ============================================================
% SLIDE 7: Early results with entropy
% ============================================================
\begin{frame}{Entropy Weighting: Helps Early, Hurts Late}

\textit{\scriptsize Synthetic benchmark: GPT-2 124M + LoRA, sentiment reward, 150 episodes}

\begin{table}
\begin{tabular}{lccc}
\toprule
Method & R(Q4) at ep 100 & R(Q4) at ep 150 & Trend \\
\midrule
Entropy-weighted PPO & \textbf{0.872} & 0.797 & \textcolor{bad}{$\downarrow$} \\
PPO baseline & 0.791 & \textbf{0.850} & \textcolor{good}{$\uparrow$} \\
\bottomrule
\end{tabular}
\end{table}

\vspace{0.3cm}
\textbf{+10\% ahead at ep 100} $\longrightarrow$ \textbf{surpassed by ep 150}

\vspace{0.3cm}
The weights stay fixed, but the model changes. Tokens that were uncertain early may no longer be the right ones to focus on --- the fixed weights become stale and the error accumulates.

\end{frame}

% ============================================================
% SLIDE 8: Why degradation happens
% ============================================================
\begin{frame}{Why Do Fixed Weights Degrade?}

\textbf{The gradient estimator with entropy weights $w(t)$ is biased:}
\[
g_w = \mathbb{E}_t\bigl[w(t) \cdot \nabla\log\pi(a_t|s_t) \cdot \hat{A}_t\bigr], \qquad \text{Bias} = \text{Cov}(w, f) \neq 0
\]

This bias is small per step, but it \textbf{accumulates} over hundreds of PPO updates.

\vspace{0.3cm}
\begin{columns}
\begin{column}{0.48\textwidth}
\textbf{Early in fine-tuning:}
\begin{itemize}
    \item Gradient variance is high (policy still adapting)
    \item Entropy weights reduce this variance by focusing on informative tokens
    \item Net effect: \textcolor{good}{faster learning}
\end{itemize}
\end{column}
\begin{column}{0.48\textwidth}
\textbf{Later in fine-tuning:}
\begin{itemize}
    \item Gradient variance is lower (policy stabilizing)
    \item But accumulated bias from fixed weights distorts the gradient direction
    \item Net effect: \textcolor{bad}{wrong updates, reward declines}
\end{itemize}
\end{column}
\end{columns}

\vspace{0.3cm}
Even though the base model is pretrained and capable, the RL fine-tuning process itself has early/late dynamics --- the policy changes most in the first few hundred PPO steps.

\end{frame}

% ============================================================
% SLIDE 9: AITI
% ============================================================
\begin{frame}{Solution: Adaptive Intensity Token Importance (AITI)}

\textbf{Decay the entropy-weighting intensity as training progresses:}
\[
w_{\text{AITI}}(t) = 1 + \alert{\varepsilon(\text{step})} \cdot \bigl(H(t)_{\text{norm}} - 1\bigr)
\]

\begin{itemize}
    \item $\varepsilon = 1$: full entropy weighting (focus on uncertain tokens)
    \item $\varepsilon = 0$: uniform PPO (all tokens equal)
    \item $\varepsilon$ decays linearly over a configurable number of steps
\end{itemize}

\vspace{0.3cm}
\textbf{Analogy:} Learning rate scheduling controls \textit{step size}. AITI controls \textit{where} the gradient focuses.

\begin{table}
\small
\begin{tabular}{lcc}
\toprule
Fine-tuning phase & Learning Rate & AITI Intensity ($\varepsilon$) \\
\midrule
Early (policy adapting) & High & High (focus on uncertain tokens) \\
Late (policy stabilizing) & Low & Low (equal treatment, no bias) \\
\bottomrule
\end{tabular}
\end{table}

\end{frame}

% ============================================================
% SLIDE 10: AITI results
% ============================================================
\begin{frame}{AITI Results: Entropy Weighting That Doesn't Degrade}

\textit{\scriptsize Synthetic benchmark: GPT-2 124M + LoRA, sentiment reward, 150 episodes}
\begin{table}
\begin{tabular}{lcccc}
\toprule
Method & Reward (Q4) & KL (Q4) & KL-Efficiency & Pareto? \\
\midrule
\textbf{AITI-Entropy} & \textbf{0.855} & \textbf{0.016} & \textbf{54.1} & \textcolor{good}{\textbf{YES}} \\
PPO baseline & 0.818 & 0.040 & 20.4 & dominated \\
Entropy (fixed, no decay) & 0.815 & $-$0.611 & 1.3 & dominated \\
TI-DPO Hybrid & 0.779 & 0.024 & 32.7 & dominated \\
\bottomrule
\end{tabular}
\end{table}

\vspace{0.3cm}
\begin{itemize}
    \item AITI-Entropy \textbf{Pareto-dominates} PPO: higher reward \textit{and} 2.5$\times$ lower KL
    \item Fixed entropy degrades as predicted (negative KL = policy drifted \textit{toward} reference)
    \item The decay schedule is what makes entropy weighting viable in PPO
\end{itemize}

\end{frame}

% ============================================================
% SLIDE 11: MOAI motivation
% ============================================================
\begin{frame}{Can We Pick the Optimal $\varepsilon$ Automatically?}

AITI requires choosing a decay schedule. Can we compute the best $\varepsilon$ per step?

\vspace{0.2cm}
\textbf{Minimize MSE of the gradient estimator} (bias$^2$ + variance):
\[
\varepsilon^* = \arg\min_\varepsilon\; \bigl[\text{Bias}(\varepsilon)^2 + \text{Var}(\varepsilon)\bigr]
\]

\textbf{Closed-form solution} (MOAI = MSE-Optimal Adaptive Intensity):
\[
\varepsilon^* = \text{clip}\!\left(\frac{-C}{2\rho},\; 0,\; \varepsilon_{\text{prev}}\right)
\]

\begin{itemize}
    \item $C$, $\rho$ = covariance and variance statistics, estimated via EMA
    \item Monotone constraint: $\varepsilon$ can only decrease (no oscillation)
    \item No hyperparameter tuning --- adapts automatically to the training dynamics
\end{itemize}

\end{frame}

% ============================================================
% SLIDE 12: MOAI finding
% ============================================================
\begin{frame}{MOAI Finding: Partial Weighting Can Be Optimal}

% \textit{\scriptsize Synthetic benchmark: GPT-2 124M + LoRA, sentiment reward, 150 episodes}

\textbf{Empirical observation:} MOAI converges to $\varepsilon^* \approx 0.3$--$0.5$, not zero.

\vspace{0.2cm}
This seems to contradict AITI (which decays to 0). But the MSE decomposition explains it:

\begin{itemize}
    \item AITI decays to $\varepsilon = 0$ to \textbf{eliminate all bias} --- safe but conservative
    \item MOAI finds that a \textbf{small amount of bias is worth the variance reduction} it buys
    \item At $\varepsilon \approx 0.3$: the remaining bias$^2$ is tiny, but the variance reduction is still meaningful
\end{itemize}

\vspace{0.3cm}
\begin{table}
\begin{tabular}{lcccc}
\toprule
Method & Reward (Q4) & KL (Q4) & KL-Eff & Pareto? \\
\midrule
\textbf{MOAI-Entropy} & \textbf{0.880} & 0.056 & 15.7 & \textcolor{good}{\textbf{YES}} \\
AITI-Entropy & 0.855 & \textbf{0.016} & \textbf{54.1} & \textcolor{good}{\textbf{YES}} \\
PPO baseline & 0.818 & 0.040 & 20.4 & dominated \\
Entropy (fixed) & 0.815 & $-$0.611 & 1.3 & dominated \\
\bottomrule
\end{tabular}
\end{table}

MOAI achieves the \textbf{highest reward}; AITI achieves the \textbf{best KL-efficiency}. Both Pareto-dominate PPO.

\end{frame}


% ============================================================
% SLIDE 14: GSM8K Results
% ============================================================
\begin{frame}{GSM8K Results (1.5B Model)}

\begin{columns}
\begin{column}{0.52\textwidth}
\textbf{Test accuracy (1,319 examples, greedy):}

\begin{table}
\scriptsize
\begin{tabular}{lcc}
\toprule
Method & Test Acc & vs Base \\
\midrule
PPO baseline & 63.1\% & +8.4\% \\
MOAI-Entropy 0.80 & 63.0\% & +8.3\% \\
AITI-Entropy & 62.5\% & +7.8\% \\
Entropy (fixed) & 62.7\% & +8.0\% \\
\hdashline
Base model & 54.7\% & --- \\
\bottomrule
\end{tabular}
\end{table}

\vspace{0.2cm}
All methods improve by $\sim$8\% over the base model. The differences between methods are small ($<$1\%).

\textbf{Why?} GSM8K is relatively easy for this model --- the base already gets 55\%, leaving limited room for PPO to differentiate methods.
\end{column}
\begin{column}{0.46\textwidth}
\textbf{Batch size matters:}

\begin{center}
\vspace{0.5cm}
\fbox{\parbox{0.85\textwidth}{\centering\vspace{1.5cm}\textit{Batch size sweep figure}\\\textit{(to be updated)}\vspace{1.5cm}}}
\end{center}

\scriptsize
At bs=1, entropy methods give +4--5\% over baseline. At bs=8, the gap vanishes. Entropy weighting helps most when gradient variance is high.
\end{column}
\end{columns}

\end{frame}

% ============================================================
% SLIDE 15: MATH Results
% ============================================================
\begin{frame}{MATH-500 Results (Qwen3-4B, 500 Episodes)}

\textbf{Competition-level math} --- much harder than GSM8K.

\vspace{0.2cm}
\begin{table}
\small
\begin{tabular}{lcc}
\toprule
Method & Train Acc (ep 40) & Test Acc \\
\midrule
AITI-Entropy (decay=100) & \textbf{35.4\%} & \textit{running} \\
AITI-Entropy (residual=0.3) & 34.8\% & \textit{running} \\
MOAI-Entropy 0.95 & 34.2\% & \textit{running} \\
PPO baseline & 33.5\% & \textit{running} \\
Entropy (fixed) & 33.5\% & \textit{running} \\
AITI-Entropy (decay=200) & 33.5\% & \textit{running} \\
MOAI-Entropy 0.80 & 32.3\% & \textit{running} \\
MOAI-Entropy 0.90 & 31.1\% & \textit{running} \\
\bottomrule
\end{tabular}
\end{table}

\vspace{0.2cm}
\textbf{Early signal (ep 40/500):} AITI-Entropy (decay=100) leads at 35.4\% vs baseline 33.5\%. Training ongoing --- test results to follow.

\vspace{0.1cm}
\textit{\scriptsize Qwen3-4B-Instruct + LoRA, 7,500 train examples, MATH-500 test, entropy-only methods.}

\end{frame}

% ============================================================
% SLIDE 16: Takeaways
% ============================================================
\begin{frame}{Key Takeaways}

\begin{enumerate}
    \item \textbf{Entropy is the right importance signal for PPO}
    \begin{itemize}
        \item It identifies tokens where the model is uncertain --- decision points
        \item Free to compute, content-aware, no extra forward/backward passes
    \end{itemize}

    \vspace{0.2cm}
    \item \textbf{Fixed entropy weighting degrades over training}
    \begin{itemize}
        \item Biased gradient estimator: Cov$(w, f) \neq 0$
        \item Helps early (variance reduction) but hurts late (accumulated bias)
    \end{itemize}

    \vspace{0.2cm}
    \item \textbf{AITI and MOAI fix this with adaptive intensity}
    \begin{itemize}
        \item AITI: scheduled decay $\varepsilon: 1 \to 0$ (safe, conservative)
        \item MOAI: per-step optimal $\varepsilon^*$ via MSE minimization (higher reward, keeps some weighting)
    \end{itemize}

    \vspace{0.2cm}
    \item \textbf{Strongest in high-variance regimes}
    \begin{itemize}
        \item Small batch sizes, sparse rewards, early training
        \item Entropy weighting is variance reduction for the policy gradient
    \end{itemize}
\end{enumerate}

\end{frame}

% ============================================================
% SLIDE 17: Future Work
% ============================================================
\begin{frame}{Future Directions}

\begin{columns}
\begin{column}{0.48\textwidth}
\textbf{Scaling:}
\begin{itemize}
    \item Larger models (4B, 7B) where PPO has more headroom
    \item Longer training (1000+ episodes)
    \item Harder tasks: code generation, tool use
\end{itemize}

\vspace{0.3cm}
\textbf{Methods:}
\begin{itemize}
    \item Entropy-weighted KL penalty (not just policy loss)
    \item Combine with GRPO, process reward models
    \item Per-layer or per-head importance
\end{itemize}
\end{column}
\begin{column}{0.48\textwidth}
\textbf{Theory:}
\begin{itemize}
    \item Tighter bounds on bias-variance crossover
    \item Optimal decay beyond MSE criterion
    \item Connection to exploration in token space
\end{itemize}

\vspace{0.3cm}
\textbf{Core insight:}

\begin{center}
\fbox{\parbox{0.9\columnwidth}{\centering
\textit{Token importance is gradient variance reduction.}\\[3pt]
Like all variance reduction, it helps most when variance is the bottleneck.
}}
\end{center}
\end{column}
\end{columns}

\end{frame}

\end{document}
